---
Title: "Practical Machine Learning Project"
Author: "DataVizz"

---

## Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement-a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Load Data

In this section, load the data and the 20 cases that will be submitted to coursera.

```{r}
data <- read.csv('./pml-training.csv', header=T)
submit <- read.csv('./pml-testing.csv', header=T)
```

# Cleanup the data
Here, I remove columns full of NAs and remove features that are not in the submit set. The features containing NAs are the variance, mean and stddev within each window for each feature. Since the submit dataset has no time-dependence, these values are useless and can be disregarded. I also remove the first 7 features since they are related to the time-series or are not numeric.

```{r}
# Remove columns full of NAs.
features <- names(submit[,colSums(is.na(submit)) == 0])[8:59]
# Only use features used in submit cases.
train_in <- data[,c(features,"classe")]
validation <- submit[,c(features,"problem_id")]
```

# Data Partitioning

Since I'll be predicting classes in the testing dataset, I'll split the training data into training and testing partitions and use the pml-testing.csv as a validation sample. I'll use cross validation within the training partition to improve the model fit and then do an out-of-sample test with the testing partition.

```{r}
library(caret)
set.seed(123)
inTrain = createDataPartition(data$classe, p = 0.75, list = F)
training = train_in[inTrain,]
testing = train_in[-inTrain,]
```

# Feature Selection

Some features may be highly correlated. The PCA method mixes the final features into components that are difficult to interpret; Hence, I consider only features with correlation (<90%).

```{r}
outcome = which(names(training) == "classe")
highCorrCols = findCorrelation(abs(cor(training[,-outcome])),0.90)
highCorrFeatures = names(training)[highCorrCols]
training = training[,-highCorrCols]
outcome = which(names(training) == "classe")
```

The features with high correlation are accel_belt_z, roll_belt, accel_belt_y, accel_belt_x, gyros_arm_y, gyros_forearm_z, and gyros_dumbbell_x.

# Feature Ranking

The random forest method reduces overfitting and is good for nonlinear features. First, to see if the data is nonlinear, I use the random forest to discover the most important features. The feature plot for the 4 most important features is shown.

```{r}
library(randomForest)
fsRF = randomForest(training[,-outcome], training[,outcome], importance = T)
rfImp = data.frame(fsRF$importance)
impFeatures = order(-rfImp$MeanDecreaseGini)
inImp = createDataPartition(data$classe, p = 0.05, list = F)
featurePlot(training[inImp,impFeatures[1:4]],training$classe[inImp], plot = "pairs")
```

The important features are:

pitch_belt
yaw_belt
total_accel_belt
gyros_belt_x

## Model selection for training

Train the model using the random forest and k-nearest neighbors..

```{r}
ctrlKNN = trainControl(method = "adaptive_cv")
modelKNN = train(classe ~ ., training, method = "knn", trControl = ctrlKNN)
ctrlRF = trainControl(method = "oob")
modelRF = train(classe ~ ., training, method = "rf", ntree = 200, trControl = ctrlRF)
resultsKNN = data.frame(modelKNN$results)
resultsRF = data.frame(modelRF$results)
```

## Testing sample error

The random forest will give a larger accuracy compared to k-nearest neighbors. Here, I give the confusion matrix between the KNN and RF models to see how much they agree on the test set, then I compare each model using the test set outcomes.

```{r}
fitKNN = predict(modelKNN, testing)
fitRF = predict(modelRF, testing)
```

## Conclusion

```{r}

cfmat_KNN = confusionMatrix(fitKNN, testing$classe)

cfmat_KNN

cfmat_rf = confusionMatrix(fitRF, testing$classe)

cfmat_rf

```

The random forest fit is more accurate than the k-nearest neighbors method with 99% accuracy.


## Generating Files to submit as answers for the Assignment:

Finally, using the provided Test Set out-of-sample error.

For Random Forests we use the following formula, which yielded a much better prediction in in-sample:

```{r}
submit_pred <- predict(modelRF, validation)

submit_pred
```










